Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-23 ResponseLetter.tex

2. Do a search for multi-type branching and make sure you consistently use the hyphen or not.

\textcolor{blue}{%change 31
We have made the change to use the hyphen in all mentions of ``multi-type branching'' such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).
}

3. You identify that at least in certain scenarios aneuploidy plays an important role in the evolutionary rescue of tumor cell populations following chemotherapy, at least with one type of drug, and maybe limited to small primary or secondary tumors. But, we never really know the extent that this scenario is applicable. Is it specific to this particular chemotherapy, small tumors, only certain types of cancer? In conjunction with this, there are places in the manuscript where you say things like ``for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step).'' Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue. Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care. If the answer is ``I don't know how applicable this finding is or what the broader implications might be,'' you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically. It's ok to not know for sure, but you shouldn't pass the opportunity to talk about it in limits and future research.

\textcolor{blue}{
Thank you for this remark, you are correct to note that this statement is in opposition to some of our results. We have removed this statement and instead, we now write (line 523): ``The `stationary' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting. Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig. 4A, inset), it could be biologically significant for some tumors under specific drug therapies. 
Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
We also added the following sentences to the Discussion.
In line 468, emphasis on new text): ``Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$. \emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}'' \\
And in line 493: ``AAs an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug. 
It remains to be seen which tumor type and drug combinations produce stationary aneuploidies.''\\
The following sentences were already in the Discussion (line 476): ``In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation. 
Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
Together, this increases the cumulative number of mutants that arise.'' and in line 481: ``We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance.''
}

@@

\textcolor{blue}{Thank you for the positive review of our manuscript.}

Major points

- The importance of time-resolved measurements of clonal dynamics have been recognized and more such datasets are becoming available (e.g., PMID: 34163070, PMID: 38987605). The simulation studies in this manuscript use a parameter regime largely based on a melanoma cell-line (A375) and treatment with a specific BRAF-inhibitor (vemurafenib). The relevance of the paper would be improved if consistent results can be shown in other datasets, including the citations above.

\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B.\\
We have aldo added the following text to the Discussion (line 493): ``As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug. 
It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
Comparing the probability of evolutionary rescue between different tumors (Figure 3B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones, because smaller tumors have higher resuce probability. Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.'' % change A
}% change A

@@

Results:
* Figure 7 - Add a numerical solution to equation (D3)? This should make it clearer that the non-monotonicity of the black curve is an artefact of the simplifying assumptions in (D4) and not a problem with the approach.

\textcolor{blue}{We have added the green line for eq. D3 to Figure 7 as suggested. It is very close to the blue line for eq. D4.}

* By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)? Could this be briefly clarified?

\textcolor{blue}{In the first term of (D3) we have: 
\begin{align*}
\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
\end{align*}
which is significantly smaller then the second term of (D3):
\begin{align*}
v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\Delta_m-\Delta_s}
\end{align*}
} % change 44
\\
\textcolor{red}{I don't understand our answer here. If aneuploids are common and pretty good, can't they still matter? And I don't understand what we're saying about ``large $\tau_a^r$''---large compared to what? Aren't we describing the regime of relatively small $\tau_a^r$ here? }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-01-23 ms.tex
minor formatting

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
This equation is equivalent to eq.~A3 of \citet{iwasa2004evolutionary} and to eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
our ``tolerant'', ``stationary'', and ``resistant'' scenarios correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively. %change 57
These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
Our parameterization for two types of cancers suggests that melanoma and for TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$). 
In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.

@@

\section{Recurrence time}\label{sec:appendix_recurrence_time}
We define the proliferation time $\tau_a^p$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$. The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
\begin{equation*}
r_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
\end{equation*}
where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \Cref{eq:N_a}.

@@

The second scenario is when the sensitive population produces a large number of rescue lineages in a short period of time. This is given by the condition:
\begin{align*}
N\gg N_a^*.
\end{align*}
As a result, the recurrence time is obtained by solving the following system of ODEs:
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation}
We obtain $\tau_a^r$ by solving \eqref{m_t} for the time at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}
This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as: 
\begin{equation*}
1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\Delta_m - \Delta_s},
\end{equation*}
which has solution:
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}.
\end{equation}
We observe that the terms given by the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N \gg N_m^* > N_a^*$. %change 9

Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \Cref{meanproliferationtime} we change:
\begin{align}\label{meanproliferationtime2}
\tau_a^{r,M}&\approx\tau_a+\frac{\log p_mM}{\Delta_m},
\end{align}
and \Cref{eq:t2det} becomes:
\begin{equation}\label{eq:t2det2}
\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_s\right)}{v\lambda_sN},
\end{equation}
which we plot in \Cref{RecurrencePlot} and observe that our approximations are in agreement with simulations.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
This equation is equivalent to eq.~A3 of \citet{iwasa2004evolutionary} and to eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
our ``tolerant'', ``stationary'', and ``resistant'' scenarios correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively. %change 57
These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
Our parameterization for two types of cancers suggests that melanoma and TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$). 
In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.

@@

\section{Recurrence time}\label{sec:appendix_recurrence_time}
We define the proliferation time $\tau_a^p$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$. The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
\begin{equation*}
r_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
\end{equation*}
where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \cref{eq:N_a}.

@@

The second scenario is when the sensitive population produces a large number of rescue lineages in a short period of time. This is given by the condition:
\begin{align*}
N\gg N_a^*.
\end{align*}
As a result, the recurrence time is obtained by solving the following system of ODEs:
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}
This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as: 
\begin{equation*}
1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\Delta_m - \Delta_s},
\end{equation*}
which has solution:
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}.
\end{equation}
We observe that the terms given by the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N \gg N_m^* > N_a^*$. %change 9

Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \cref{meanproliferationtime} we change:
\begin{align}\label{meanproliferationtime2}
\tau_a^{r,M}&\approx\tau_a+\frac{\log p_mM}{\Delta_m},
\end{align}
and \Cref{eq:t2det} becomes:
\begin{equation}\label{eq:t2det2}
\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_s\right)}{v\lambda_sN},
\end{equation}
which we plot in \Cref{RecurrencePlot} and observe that our approximations are in agreement with simulations.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
